A Stopped Negative Binomial Distribution

### 2016-03-30 stat-and-prob-letters/snb-paper.tex
Integrating Dan's changes.

\begin{abstract}
We introduce a discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed to observe either $s$ successes 
or $t$ failures. The closed-form expression for the distribution as well as 
its characteristics are derived and properties of the distribution are explored.
\end{abstract}

@@

Consider a prototypical Phase II single-arm clinical trial in which 12 patients
are enrolled and treated. If two or more respond then we reject the null
hypothesis that the treatment is not effective and the trial is deemed
successful. If fewer than two respond then the null is not rejected and the 
trial is terminated.

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, as in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us to reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size is between two and 12.

Assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables that stops when either a
specified number of successes or failures is reached. In the
previous example suppose two successes were reached after enrolling 10
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial.

@@

\begin{table}[t!]
\caption{Stopped Negative Binomial Distribution Characteristics}
\label{tab:snb}
\begin{center}
\begin{tabular}{|c|l|} \hline
Notation & SNB($p$, $s$, $t$) \\ \hline
Parameters & $p$ the success probability ($q = 1-p$) \\
           & $s$ the number of successes before stopping \\
           & $t$ the number of failures before stopping \\ \hline
Support & min$(s,t) \leq k \leq s+t-1$  \\ \hline
PMF & ${k-1 \choose s-1} p^s (1-p)^{k-s} + {k-1 \choose t-1} (1-p)^s p^{k-t}$\\ \hline
CDF & $2 - \mathcal{I}_{1-p}(k+1, s) - \mathcal{I}_{p}(k+1, t)$\\ 
    & where $\mathcal{I}$ is the regularized incomplete beta function.\\ \hline
Mean & $\frac{s}{p} \mathcal{I}_p(s,t) + \frac{q^{t-1} p^{s-1}}{B(s,t)} +
  \frac{t}{q} \mathcal{I}_q(t,s) + \frac{q^{s-1}p^{t-1}}{B(s,t)}$\\ 
  & where $B$ is the beta function \\ \hline
MGF & $\left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s) $\\ \hline
\end{tabular}
\end{center}
\end{table}

In this work we derive the distribution of the number of enrollees needed
for either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB) and some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
The rest of this paper derives these results
and explores properties of the distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
its relation to other distributions. Section 2 derives the distribution
based on a defined Bernoulli process and gives some basic properties.
Section 3 provides a connection to the Binomial tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a successful outcome.  Let $s$ and $t$ be
positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ successes {\em or} $\,t\,$ failures. That is, the SNB distribution
of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{E}\  I_{\{Y=k\}} = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is one if $f$ is true and zero otherwise and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability

>>>>

\begin{abstract}
We introduce a discrete distribution suggested by curtailed
sampling rules common in early-stage clinical trials. We derive the
distribution of the smallest number of independent and identically
distributed Bernoulli trials needed to observe either $s$ successes 
or $t$ failures. This report provides a closed-form expression for the 
mass function and illustrates limiting approximations.
\end{abstract}

@@

Consider a prototypical Phase II single-arm clinical trial in which 12 patients
are enrolled and treated. The binomial probability of success is $p=0.045$ for
each patient under the null hypothesis that treatment is no effective.
If two or more patients respond to the treatment then we reject this null
hypothesis, and the trial is deemed successful at significance level of $0.1$. 
If fewer than two respond then the null is not rejected.

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, as in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us to reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size for this trial is between two and 12.

Let us assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables stopping when either a
specified number of successes or failures has been reached.

A hypothetical sample path is illustrated in Fig.~\ref{fig:kane_viz}.
The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial. In this case two successes were reached after
enrolling 10 patients: one in the third outcome and on in the tenth.

@@

\begin{table}[t!]
\caption{Characteristics of the Stopped Negative Binomial Distribution}
\label{tab:snb}
\begin{center}
\begin{tabular}{|c|l|} \hline
Parameters & $p$ the success probability ($0\leq p \leq 1$; $q = 1-p$) \\
           & $s$ the number of successes before stopping ($s=1, 2, ...$)\\
           & $t$ the number of failures before stopping ($t=1, 2, ...$)\\ \hline
Support & min$(s,t) \leq k \leq s+t-1$  \\ \hline
$\mathbb{P}[Y=k]$ & ${k-1 \choose s-1} p^s (1-p)^{k-s} + {k-1 \choose t-1} (1-p)^s p^{k-t}$\\ \hline
$\mathbb{P}[Y \leq k]$ & $2 - \mathcal{I}_{q}(k+1, s) - \mathcal{I}_{p}(k+1, t)$\\ 
    & where $\mathcal{I}$ is the regularized incomplete beta function.\\ \hline
Mean & $\frac{s}{p} \mathcal{I}_p(s,t) + \frac{p^{s-1} q^{t-1}}{B(s,t)} +
  \frac{t}{q} \mathcal{I}_q(t,s) + \frac{p^{t-1} q^{s-1}}{B(s,t)}$\\ 
  & where $B$ is the beta function \\ \hline
MGF & $\left(\frac{p e^x}{1 - qe^x}\right)^s 
  \mathcal{I}_{1-qe^x} (s, t) + \left(\frac{qe^x}{1-pe^x}\right)^t 
  \mathcal{I}_{1-pe^x}(t, s) $\\ \hline
\end{tabular}
\end{center}
\end{table}

In this work we derive the distribution of the number of enrollees needed
for either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB) and some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
The rest of this paper derives these results
and explores properties of the distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
its relation to other distributions. Section 2 derives the distribution
based on a defined Bernoulli process and gives some basic properties.
Section 3 provides a connection to the Binomial tail probability.
Section 4 derives the posterior distribution using a Beta prior distribution
for $p$. Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically
distributed, Bernoulli random variables with $\mathbb{P}[b_i=1]=p$ and
$\mathbb{P}[b_i = 0] = 1-p$, for
probability parameter $0\leq p \leq 1$. In the clinical trial setting
$\,b_i = 1$ corresponds to a successful patient outcome following treatment.  
Let $s$ and $t$ be positive integers.  Define the SNB random
variable $Y$ as the smallest
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either}
$\,s\,$ successes {\em or} $\,t\,$ failures. That is, the SNB distribution
of $Y$ is the smallest integer such that either
$\sum_i^Y b_i = s$ or $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}.
\end{equation*}
The probability mass function is
\begin{equation} \label{eqn:pmf}
\mathbb{P} [Y=k] = S(k, p, s) \ I_{\{s \leq k\}} + 
  S(k, 1-p, t) \ I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is one if $f$ is true and zero otherwise, and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
is the negative binomial probability mass.
